---
title: "Regression"
author: "Dr. Irene Vrbik"
subtitle: "STAT 205: Introduction to Mathematical Statistics"
institute: "University of British Columbia Okanagan"
format:
  revealjs:
    html-math-method: mathjax
    chalkboard: true
    slide-number: true
revealjs-plugins:
  - quiz
filters:
  - webr
webr:
  packages: ["tibble", "ggplot2"]
bibliography: ../references.bib  
---

```{r}
#| include: false

library(knitr)
library(kableExtra)
library(tidyverse)
library(ggfortify)
library(janitor)
library(broom)

theme_set(theme_bw())
```

## Introduction

:::: {style="font-size: 90%;"}
::: incremental
- In this lecture we will introduce Simple Linear Regression (SLR) to model the relationship between two numerical variables.

- We will quantify the relationship between variables using popular metrics like the Correlation Coefficient and $R^2$ value.

- Statistical inference for this model will be discussed, and the critical distinction between inference for mean response and inference for the outcome will be clarified.

- We will also introduce a basic understanding of the multiple regression model.
:::
::::

## Recall: Scatterplots

```{webr-r}
#| context: setup
#| autorun: true

theme_set(theme_bw())
```

```{webr-r}
#| autorun: false

set.seed(2025); n = 8
atib <- tibble(x = rnorm(8), y = 2 + 4*x + rnorm(n, sd = 2))
atib %>% ggplot(aes(x = x, y = y)) + 
  geom_point(colour="red", size = 3) 
```

## Line of Best fit

```{r}
#| echo: false

set.seed(2025); n = 8
atib <- tibble(x = rnorm(8), y = 2 + 4*x + rnorm(n, sd = 2))
```

```{r}
#| fig-width: 7
#| fig-height: 5
#| fig-align: "center"


# Fit a linear model
model <- lm(y ~ x, data = atib)

# Plot data with regression line only
ggplot(atib, aes(x = x, y = y)) +
  geom_point(color = "red", size = 3) +                     # Data points
  geom_smooth(method = "lm", se = FALSE, color = "blue") + # Regression line
  theme_bw() +                                              # Match scatterplot theme
  labs(title = NULL)                                        # No title

```

## Visualization of Residuals

```{r}
#| fig-width: 7
#| fig-height: 5
#| fig-align: "center"


# Add fitted values and residuals to the data
atib <- atib %>%
  mutate(fitted = model$fitted.values,
         resid = model$residuals,
         label = paste0("epsilon[", row_number(), "]"))

ggplot(atib, aes(x = x, y = y)) +
  geom_point(color = "red", size = 2) +                         # Original data points
  geom_smooth(method = "lm", se = FALSE, color = "blue") +     # Regression line
  geom_segment(aes(xend = x, yend = fitted),                   # Residual lines
               linetype = "dashed", color = "gray") +
  geom_text(aes(label = label, y = y + 0.5), parse = TRUE) +   # Epsilon labels
  theme_bw() +                                                 # Match theme
  labs(title = NULL)                                           # Remove title to match scatterplot

```

## What is regression

::: {style="font-size: 90%;"}
Regression analysis is a tool to investigate how two or more variables are related.

> *"The general objective of a regression analysis is to determine the relationship between two (or more) variables so that we can gain information about one of them through knowing values of the other(s)."*

<p style="text-align: right;">

Source: Ch 12 of @devore2021modern

</p>
:::

## Variable Types

We want to see how a specific **response variable**, $Y$ is affected by one or more **explanatory** **variables** $X_1, \dots, X_p$

::: incremental
- $Y$ is also known as the **dependent** **variable** or **output variable.** This variable is continuous (i.e. numeric data type)

- $X_i$'s are also known as **independent/predictor variables** or **features.** These may be continuous or categorical; however, we only consider continuous predictors in this course.
:::

##  {.quiz-question}

:::: {.callout-note icon="false"}
## Example

::: exr-height
One may wish to use a person's height, gender and race to predict a person's weight. What is/are the response variable(s)?
:::
::::

- Height
- Gender
- Race
- [Weight]{.correct}

##  {.quiz-question}

:::: {.callout-note icon="false"}
## Example

::: exr-height
One may wish to use a person's height, gender and race to predict a person's weight. What is/are the dependent variable(s)?
:::
::::

- [Height]{.correct data-explanation="This is one potential answer."}
- [Gender]{.correct data-explanation="This is one potential answer."}
- [Race]{.correct data-explanation="This is one potential answer."}
- [Weight]{data-explanation="This is the response variable. It would not make sense to use as a depedent variable as well."}

## Model

$$\begin{align}
Y &= \underbrace{f(X_1, \dots, X_p)}_{\text{deterministic}} + \underbrace{\epsilon}_{\text{random}} \\
&\phantom{= \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \dots \beta_p X_p + \epsilon}
\end{align}$$

## Model

$$\begin{align}
Y &= \underbrace{f(X_1, \dots, X_p)}_{\text{deterministic}} + \underbrace{\epsilon}_{\text{random}}\\
&= \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \dots \beta_p X_p + \epsilon
\end{align}$$

. . .

In a **linear regression model**, the function $f(X_1, \dots, X_p)$ is a *linear* combination of the explanatory variables.

## Model

$$\begin{align}
\overset{\overset{\text{response}}{\downarrow}}{Y} &= \beta_0 + \beta_1 \underset{\underset{\text{explanatory}}{\uparrow}}{X_1} + \beta_2 \underset{\underset{\text{explanatory}}{\uparrow}}{X_2} + \dots \beta_p \underset{\underset{\text{explanatory}}{\uparrow}}{X_p} + \epsilon
\end{align}$$

:::: {style="font-size: 90%;"}
::: incremental
- $\beta_0$ is the $y$-intercept

- $\beta_i$ are the regression coefficients, and

- $\epsilon$ are the errors that are assumed $N(0, \sigma)$
:::
::::

::: notes
In statistics, we can describe how variables are related using a mathematical function.

The function along with other assumptions is called a model.

There are many models we can consider. In this class, we will focus on linear models, particularly, when there is only one predictor variable.

We refer to this model as the **simple linear regression model**. (Penn STat 500)
:::

::: notes
as well as modeling numerical response variables using a numerical or categorical explanatory variable.
:::

## Simple Linear Regression

- When $p = 1$, namely, we are only considering one explanatory variable, our model is called *"simple" linear regression (SLR).*
- When $p>1$ we fall into *multiple linear regression (MLR).*

<!-- - Let us first consider the "simple" case: using a person's height to predict the person's weight. -->

## SLR model

The SLR model reduces to an equation for of a line:

$$Y = \beta_0 + \beta_1 X$$

where:

- $\beta_0$ is the $y$-intercept, and

- $\beta_1$ is the slope

## Simulation: Part 1 {.scrollable}

```{webr-r}
beta0 = __ # pick a y-intercept
beta1 = __ # pick a slope
```

```{webr-r}
n = 20 # number of points 
x = sample(100, n)
y = beta0 + beta1*x
plot(x, y)
```

::: notes
This is only the deterministic portion
:::

## Comment

- We have only code the *deterministic* portion of this model
- In other words, we are missing *random* component
- Let's add some random noise ...

## Simulation: Part II {.scrollable}

::: callout-warning
Change `sd` as needed
:::

```{webr-r}
beta0 = 3; beta1 = 4; n = 20; x = sample(100, n)
eps = rnorm(n, mean = 0, sd = 15) # random noise 
y = beta0 + beta1*x + eps         # linear model with noise
plot(x, y)                        # plot the data
```

## 

```{r}
#| echo: false
beta0 = 3; beta1 = 4; n = 20; 
```

::: panel-tabset
### $X$s

Create `n` values for `x`, for example using:

```{r}
#| echo: true
x = runif(n, min = -2, max = 10) # or runif() / rnorm()
```

```{r}
#| echo: false

eps = rnorm(n, mean = 0, sd = 3) # random noise 
y = beta0 + beta1*x + eps         # linear model with noise
df <- data.frame(x = x, y = y)

```

```{r}
#| meesage: false

g = ggplot(df, aes(x = x, y = y)) +
  geom_point(color = NA) +  # add the points
  geom_abline(intercept = beta0, slope = beta1, color = "steelblue", linewidth = 0, color = NA) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  geom_vline(xintercept = 0, linewidth = 0.3) +
  scale_x_continuous(breaks = seq(0, 100, 25)) +
  labs(x = "X", y = "Y") +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_blank(),
    axis.ticks = element_line(),
    plot.margin = margin(20, 20, 20, 20)
  )

# Add both point layers
g + 
  geom_point(data = df, aes(x = x, y = 0), color = "firebrick", size = 4, shape = 21, fill = "firebrick", alpha = 0.5)


```

### Parameters

For $\beta_0$ = `beta0` = `r beta0`, $\beta_1$ = `beta1` = `r beta1`, we have

$$Y = `r beta0` + `r beta1` \cdot X$$

```{r}
g = ggplot(df, aes(x = x, y = y)) +
  geom_point(color = NA) +  # add the points
  geom_abline(intercept = beta0, slope = beta1, color = "steelblue", linewidth = 1) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  geom_vline(xintercept = 0, linewidth = 0.3) +
  scale_x_continuous(breaks = seq(0, 100, 25)) +
  labs(x = "X", y = "Y") +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_blank(),
    axis.ticks = element_line(),
    plot.margin = margin(20, 20, 20, 20)
  )

# Add both point layers
g + 
  geom_point(data = df, aes(x = x, y = 0), color = "firebrick", size = 4, shape = 21, fill = "firebrick", alpha = 0.5)

```

### Deterministic

```{r}
# Add both point layers
g + 
  geom_point(data = df, aes(x = x, y = beta0 + beta1*x), color = "firebrick", size = 4, shape = 21, fill = "firebrick", alpha = 0.5)
```

### Noise

```{r}
# Add both point layers
g + 
  geom_point(data = df, aes(x = x, y = y), color = "firebrick", size = 4, shape = 21, fill = "firebrick", alpha = 0.5)
```
:::

## References
