---
title: "An Activity in R Introducing Visual"
subtitle: "Run R code to visualize data of Pixar films"
author: "Kyran Cupido"
institute: "St. Francis Xavier University"
execute: 
  echo: true
format:
    html:
        code-fold: true
        toc: false
        mathjax: "local"
filters:
  - webr
webr:
  packages: ["pixarfilms"]
smaller: false
---

This script will provide a quick demonstration of summarizing data visually.


## Pixar Data

```{webr-r}
#| context: setup
#| autorun: true

library(dplyr)
library(pixarfilms)

pixar<-pixar_films |> 
  full_join(box_office,by="film") |> 
  full_join(public_response,by="film") |> 
  select(film, release_date, run_time, rotten_tomatoes, box_office_worldwide,  budget ) |> 
  na.omit()
  
attach(pixar)
```


```{r, include=F}

library(pixarfilms)
library(dplyr)
library(knitr)
library(kableExtra)

pixar<-pixar_films |> 
  full_join(box_office,by="film") |> 
  full_join(public_response,by="film") |> 
  select(film, release_date, run_time, rotten_tomatoes, box_office_worldwide,  budget ) |> 
  na.omit()

```

The data we will be looking at contains information about Pixar films released prior to 2021, as provided by Wikipedia. The following variables were collected for each film:

- `film`: The name of the film.
- `release_date`: The date the film premiered.
- `run_time`: The length of the film in minutes.
- `rotten_tomatoes`: Score from the review-aggregation website Rotten Tomatoes; scored out of 100.
- `box_office_worldwide`: Box office gross amount in U.S. dollars worldwide.
- `budget`: Movie budget in U.S. dollars.

The data can be found in the following table:

```{r, echo=F}
pixar |> 
kable(format.args = list(big.mark = ",", scientific = FALSE)) |> 
  kable_styling("striped", full_width = T) |> 
  scroll_box(width = "100%", height = "300px")
```

### Descriptive Statistics

The following functions are useful for summarizing **numerical data**: `mean`, `median`, `min`, `max`, `range`, `sd`, `var`, `fivenum`

Try them here:

```{webr-r}
#| autorun: false
mean(run_time)
```


### Basic Plots

There is a generic `plot` function in R which is useful for creating quick visualizations of a data set. If you pass one numeric variable into this function, a quick plot will be made by creating a point for each observation in the order they occur in the data frame (index). Passing two numeric variables into this function will create a scatterplot, with the first variable corresponding to the x-axis and the second corresponding to the y-axis.

```{webr-r}
#| autorun: false
plot(rotten_tomatoes, box_office_worldwide)
```


There are some arguments we can add to the `plot` function to make our visualizations more informative:

- main = "Title of the Graph"
- xlab = "x-axis label"
- ylab = "y-axis label"
- xlim = c(0,100) (changing the values on the x-axis)
- ylim = c(0,100) (changing the values on the y-axis)
- col = "colour" (full list of colours can be found [**here**](https://r-charts.com/colors/))
- pch = number (list of potential shapes for points can be found [**here**](https://r-charts.com/base-r/pch-symbols/))

We can even add text to our plots by incorporating the [**text function**](https://www.rdocumentation.org/packages/graphics/versions/3.6.2/topics/text).


### Barplots 

To draw a basic bar graph, we can simply use the `barplot` function and specify which variable it is that we want to represent the height of the bars. For example, we can use the following code to quickly plot the Rotten Tomatoes score for each of the Pixar movies:

```{webr-r}
#| autorun: false

barplot(rotten_tomatoes)

```

There are some arguments we can add to the `barplot` function to make our bar graph more informative.


- horiz = T (draw the bars horizontally) 
- name = variable to label the bars
- las = 2 (rotates the labels)
- cex.names = 0.5 (change the labels font size)



### Histograms 

We can create histograms of the data  using the `hist` function. For example, we can use the following code to quickly create a histogram of the Rotten Tomatoes scores for the Pixar movies:

```{webr-r}
#| autorun: false

hist(rotten_tomatoes)

```

Arguments we can add to the `hist` function to make our graph more informative.

- breaks = number (number of bars displayed)
- labels = T (add numbers to bars)


### Boxplots 

The `boxplot` function can be used to (unsurprisingly) create boxplots. Keeping with our Rotten Tomatoes examples:

```{webr-r}
#| autorun: false

boxplot(rotten_tomatoes)

```


Arguments we can add to the `boxplot` function to make our graph more informative.

- horizontal = T (draw the boxplot horizontally) 

